Integer Programming Formulations for the Shared Multicast Tree Problem

### 2018-12-28 smt/sec3-sbtsmt.tex
mainly sect 7 rewritten

In more general terms, let $T\subseteq G$ be a multicast tree, and let $i\in V_T$ be a node with at least two neighbors in $T$.
For multicasting a message, the power to be assigned to node $i$ is either $p_{ii_1}$ or $p_{ii_2}$, where $i_1$ and $i_2$ are neighbors of $i$ such that
$p_{ii_1}>p_{ii_2}>p_{ij}$ for all other neighbors $j\in V_T\setminus\{i_1,i_2\}$.
If the path from the message source to $i$ intersects $i_1$, then power $p_{ii_2}$ is sufficient.
Otherwise, the power $p_{ii_1}$ is required.
At a leaf $i$ with neighbor $i_1$ in $T$, the power assignment is $p_{ii_1}$ if $i$ also is the message source, and zero otherwise.

>>>>

In more general terms, let $T\subseteq G$ be a multicast tree, and let $i\in V_T$ be a node with at least two neighbors in $T$.
For multicasting a message, the power to be assigned to node $i$ is either $p_{ii_1}$ or $p_{ii_2}$, where $i_1$ and $i_2$ are neighbors of $i$ such that
$p_{ii_1}>p_{ii_2}>p_{ij}$ for all other neighbors $j\neq i_1,i_2$ of $i$ in $V_T$.
If the path from the message source to $i$ intersects $i_1$, then power $p_{ii_2}$ is sufficient.
Otherwise, the power $p_{ii_1}$ is required.
At a leaf $i$ with neighbor $i_1$ in $T$, the power assignment is $p_{ii_1}$ if $i$ also is the message source, and zero otherwise.

### 2018-12-28 smt/sec7-heur.tex
mainly sect 7 rewritten

\textcolor{blue}{\section{Computing near-optimal solutions} \label{sec:heur}}

\textcolor{blue}{
Instances of size that prohibits computation of the optimal solution should be approached by some fast, but possibly inexact method.
Imposing a time bound on an IP-solver, and applying it to any of the models introduced in Section \ref{sec:ILP}, readily gives such a method.
Alternatively, a heuristic method dedicated to SMT can be developed.}

\textcolor{blue}{
To the best of our knowledge, the scientific literature offers few heuristic methods for SMT besides those studied by \citet{ivanova16isco}.
However, recent progress by \citet{pajor18} on heuristic methods for the related minimum Steiner tree problem can smoothly be exploited in order to generate analogous methods for SMT.
The main algorithmic components in the most basic heuristic are a construction algorithm, a local improvement algorithm,
and an algorithm for merging two Steiner trees in order to create a better one.
Construction and local improvement are applied in order to generate Steiner trees of presumably low cost.
If the cost is sufficiently low, the tree is stored in an ordered set $\mathcal{T}$ of at most $\gamma$ distinct trees.
Diversity among the trees is achieved by perturbing the edge costs randomly before construction and local improvement are applied.
After each tree generation, the new tree is merged with a randomly chosen tree from $\mathcal{T}$, 
and the resulting tree is kept in $\mathcal{T}$ if its cost is smaller than some other tree already in $\mathcal{T}$.
If also $|\mathcal{T}|=\gamma$, the most expensive tree $T_{\gamma}$ is removed from $\mathcal{T}$.
}

\textcolor{blue}{
Adapting the method by \citet{pajor18} to our problem is straightforward,
given the recently published construction and local improvement algorithms for SMT \citep{ivanova16isco}.
Let \textit{SMTHeur}$(G,D,p)$ denote the output produced when $(G,D,p)$ is input to the construction algorithm,
and the constructed tree is input to the local improvement method.
The adapted method is then described as follows, where $\mathcal{T}=\left(T_1,\ldots,T_{|\mathcal{T}|}\right)$
and $c\left(T_1\right)\leq\dots\leq c\left(T_{|\mathcal{T}|}\right)$:}

\begin{algorithm}
\textcolor{blue}{
\While{the time assigned to the heuristic is not expired} {
  \textbf{for} $\{i,j\}\in E$ \textbf{do} $\hat{p}_{ij}\leftarrow\text{random number uniformly distributed on
  $\left[\frac{1}{2}p_{ij},\frac{3}{2}p_{ij}\right]$}$\;
  $T\leftarrow\text{\textit{SMTHeur}$(G,D,\hat{p})$}$\;
  \textbf{if} $|\mathcal{T}|<\gamma$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\;
  \Else {
    \textbf{if} $c(T)<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\;
    $r\leftarrow\text{random number uniformly distributed on $\{1,\ldots,\gamma\}$}$\;
    \textbf{for} $\{i,j\}\in E_T\cap E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow p_{ij}$\;
    \For{$\{i,j\}\in \left(E_T\cup E_{T_r}\right)\setminus \left(E_T\cap E_{T_r}\right)$}{
      $\nu\leftarrow\text{random number uniformly distributed on $\{100,\ldots,500\}$}$\;
      $\tilde{p}_{ij}\leftarrow \nu p_{ij}$\;
    }
    \textbf{for} $\{i,j\}\in E\setminus E_T\setminus E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow 1000p_{ij}$\;
    $\tilde{T}\leftarrow\text{\textit{SMTHeur}$(G,D,\tilde{p})$}$\;
    \textbf{if} $c(\tilde{T})<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{\tilde{T}\}$\;
  }
}
\Return $T_1$
}

\textcolor{blue}{
\caption{Outline of the heuristic method}
\label{alg:meta}
}
\end{algorithm}

\textcolor{blue}{
Following \citet{pajor18}, the perturbed cost $\hat{p}_{ij}$ of edge $\{i,j\}$ is put equal to a random number with expected value $p_{ij}$.
Further, in the merge process, the adjusted weights $\tilde{p}_{ij}$ are determined in order to 
promote edges that appear in both $T$ and the randomly selected tree $T_{\gamma}$ in $\mathcal{T}$. 
The cost of edges appearing in exactly one of the trees are thus magnified.
For the purpose of diversification, the factor by which the cost is multiplied is chosen randomly.
To discourage the appearance of edges outside both $T$ and $T_r$ in the merge $\tilde{T}$,
their cost is magnified by a factor dominating the factors drawn for edges that occur in one of the trees.
}

\textcolor{blue}{
A few minor simplification over the approach taken by \citet{pajor18} are made in Alg.\ \ref{alg:meta}:
Rather than selecting the tree to leave $\mathcal{T}$ randomly, we consistently remove $T_{\gamma}$ when a new tree is added.
Also, for each new tree $T$ constructed, Alg.\ \ref{alg:meta} makes only one attempt to merge it with some $T_r$ stored in $\mathcal{T}$.
This contrasts the method in the cited work, where $\tilde{T}$ is chosen as the best tree found in a sequence of merges.
} \\

>>>>

\section{Computing near-optimal solutions} \label{sec:heur}

Instances of size that prohibits computation of the optimal solution should be approached by some fast, but possibly inexact method.
Imposing a time bound on an IP-solver, and applying it to any of the models introduced in Section \ref{sec:ILP}, readily gives such a method.
Alternatively, a heuristic method dedicated to SMT can be developed.

To the best of our knowledge, the scientific literature offers few heuristic methods for SMT besides those studied by \citet{ivanova16isco}.
However, recent progress by \citet{pajor18} on heuristic methods for the related minimum Steiner tree problem can smoothly be exploited in order to generate analogous methods for SMT.
The main algorithmic components in the most basic heuristic are a construction algorithm, a local improvement algorithm,
and an algorithm for merging two Steiner trees.

A set $\mathcal{T}=\left(T_1,\ldots,T_{|\mathcal{T}|}\right)$ of at most $\gamma$ distinct Steiner trees,
ordered by non-decreasing costs, is maintained throughout the heuristic \citep{pajor18}.
Construction and local improvement are applied in order to generate a new tree $T$ of presumably low cost.
Diversity in $\mathcal{T}$ is achieved by perturbing the edge costs randomly before construction and local improvement are applied.
If $\left|\mathcal{T}\right|<\gamma$, $T$ is included in $\mathcal{T}$.
If $\left|\mathcal{T}\right|=\gamma$, and the cost of $T$ is below the cost of $T_{\gamma}$, $T_{\gamma}$ is replaced by $T$ in $\mathcal{T}$.
Another Steiner tree, $\tilde{T}$, is generated by merging $T$ with a randomly chosen $T_r\in\mathcal{T}$ ($r=1,\ldots,\gamma$),
and $\tilde{T}$ replaces $T_{\gamma}$ in $\mathcal{T}$ if its cost is smaller.

\begin{algorithm}
\While{the time assigned to the heuristic is not expired} {
  \textbf{for} $\{i,j\}\in E$ \textbf{do} $\hat{p}_{ij}\leftarrow\text{random number uniformly distributed on
  $\left[\frac{1}{2}p_{ij},\frac{3}{2}p_{ij}\right]$}$\;
  $T\leftarrow\text{\textit{SMTHeur}$(G,D,\hat{p})$}$\;
  \textbf{if} $|\mathcal{T}|<\gamma$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\;
  \Else {
    \textbf{if} $c(T)<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{T\}$\;
    $r\leftarrow\text{random number uniformly distributed on $\{1,\ldots,\gamma\}$}$\;
    \textbf{for} $\{i,j\}\in E_T\cap E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow p_{ij}$\;
    \For{$\{i,j\}\in \left(E_T\cup E_{T_r}\right)\setminus \left(E_T\cap E_{T_r}\right)$}{
      $\nu\leftarrow\text{random number uniformly distributed on $\{100,\ldots,500\}$}$\;
      $\tilde{p}_{ij}\leftarrow \nu p_{ij}$\;
    }
    \textbf{for} $\{i,j\}\in E\setminus E_T\setminus E_{T_r}$ \textbf{do} $\tilde{p}_{ij}\leftarrow 1000p_{ij}$\;
    $\tilde{T}\leftarrow\text{\textit{SMTHeur}$(G,D,\tilde{p})$}$\;
    \textbf{if} $c(\tilde{T})<c(T_{\gamma})$ \textbf{then} $\mathcal{T}\leftarrow \mathcal{T}\setminus\{T_{\gamma}\}$, $\mathcal{T}\leftarrow \mathcal{T}\cup \{\tilde{T}\}$\;
  }
}
\Return $T_1$

\caption{Outline of the heuristic method}
\label{alg:meta}
\end{algorithm}

Taking advantage of the recently published construction and local improvement algorithms for SMT \citep{ivanova16isco},
Alg.\ \ref{alg:meta} shows how the method by \citet{pajor18} is adapted to our problem.
The procedure \textit{SMTHeur}$(G,D,p)$ produces a multicast tree $T$ by first applying the construction algorithm to the input $(G,D,p)$,
and next applying local improvement to the constructed tree.
Following \citet{pajor18}, the perturbed cost $\hat{p}_{ij}$ of edge $\{i,j\}$ is put equal to a random number with expected value $p_{ij}$.
Further, in the merge process, adjusted weights $\tilde{p}_{ij}$ are determined in order to 
promote edges that appear in both $T$ and the randomly selected tree $T_r\in\mathcal{T}$. 
The cost of edges appearing in exactly one of the trees are thus magnified.
For the purpose of diversification, the factor by which the cost is multiplied is chosen randomly.
To discourage the appearance of edges outside both $T$ and $T_r$ in the merge $\tilde{T}$,
their cost is magnified by a factor dominating the factors drawn for edges that occur in one of the trees.

A few minor simplification over the approach taken by \citet{pajor18} are made in Alg.\ \ref{alg:meta}:
Rather than selecting the tree to leave $\mathcal{T}$ randomly, we consistently remove $T_{\gamma}$ when a new tree is added.
Also, for each new tree $T$ constructed, Alg.\ \ref{alg:meta} makes only one attempt to merge it with some $T_r\in\mathcal{T}$.
This contrasts the method in the cited work, where $\tilde{T}$ is chosen as the best tree found in a sequence of merges.
